# R plotly

[Plotly](https://plotly.com/graphing-libraries/) is a graphing library that allows you to create interactive charts and maps.
These are in the form of html files.

This works with `python`, `R`, `ggplot2`, and other languages.
You can create plots with [`plotly's`](https://plotly.com/r/) own library in `R`.
However, we'll demonstrate how to plotly plots with [`ggplot2`](https://plotly.com/ggplot2/).

You can even embed these in online books like the below chart.
Try hovering over points to get more information on them.

```{r, eval=TRUE, warning=FALSE}
library("ggplot2")
library("plotly")
library(palmerpenguins)
data(package = 'palmerpenguins')
scatter_plot <- ggplot(data = penguins, aes(x = flipper_length_mm, y = body_mass_g)) +
  geom_point(aes(color = species, 
                 shape = species),
             size = 2) +
  scale_color_manual(values = c("darkorange","darkorchid","cyan4")) 
ggplotly(scatter_plot)
```

## Jupyter notebook

As always we'll start off by creating a new `R` notebook and loading in the required libraries.

Create a new notebook called "plotly_tut", the location does not matter.

The libraries required are:

```{r, eval=FALSE}
#Libraries
library("ggplot2")
library("plotly")
library("IRdisplay")
library("htmlwidgets")
```

## Penguin data

We are going to use the same penguin data as the first plot in this chapter.
More info in the penguin data can be found on this [website](https://allisonhorst.github.io/palmerpenguins/).

First we need to load it into `R` as a __data frame__.

```{r, eval=FALSE}
#Read in penguin data
penguin_df <- read.csv("/pub14/tea/nsc206/NEOF/R_community/penguin.tsv",
                       check.names=FALSE, stringAsFactors = TRUE, sep = "\t")
#Check df
head(penguin_df)
```

## ggplot2

After having your data the next step is to create your `ggplot2` object.
We'll make a box plot of Species and sex of Penguin (x-axis) versus the Flipper length (y-axis).
We'll facet the plot by species and use sex as the x axis.
We'll save this as an object, save it as a png with `ggsave()`, and display it with `IRdisplay::display_png()`.
This is what we carried out in the main book.

```{r, eval=FALSE}
#ggplot2 boxplot
gg_boxplot <- ggplot(data= penguin_df, aes(x = sex, y = flipper_length_mm, fill = sex)) +
  geom_boxplot() +
  facet_grid(.~species)
#Save as png
ggsave(filename = "penguin_boxplot.png", plot gg_boxplot,
       device = "png", dpi = 300, units = "mm", height = 200, width = 200)
#Display png
IRdisplay::display_png(file = "penguin_boxplot.png")
```

## plotly

Let's create that plot again but using `plotly`.
Additionally, we save the `plotly` plot to an html file with `htmlwidgets::saveWidget()`.
We can then view the plot with `

```{r, eval=FALSE}
#ggplot2 boxplot
gg_boxplot <- ggplot(data= penguin_df, aes(x = sex, y = flipper_length_mm, fill = sex)) +
  geom_boxplot() +
  facet_grid(.~species)
#Create plotly object from ggplot2 object
plotly_boxplot <- plotly::ggplotly(gg_boxplot)
#Save plotly object as html
htmlwidgets::saveWidget(widget = plotly_boxplot,
                        file = "penguin_boxplot.html",
                        selfcontained=TRUE)
#Display plotly object
plotly::embed_notebook(plotly_boxplot)
```

You can also open "penguin_boxplot.html" via the jupyter navigation page to view it.