# Differential abundance (DA) analysis

We are going to carry out differential abundance analysis with [ANCOM-BC (Analysis of compositions of microbiomes with bias correction)](https://www.nature.com/articles/s41467-020-17041-7).
This is an improvement of the original ANCOM method but with added bias correction.
This bias correction aims to account for the bias caused by the difference in sampling/depth across the samples.

Differential abundance analysis aims to detect features (ASVs, taxa, etc.) that have different abundances between groups.
This can be used to detect organisms in high abundance in diseased states versus healthy states.
This helps determine what are the disease causing organisms.
However, it may also detect organisms that are able to survive in the environment rather than those that cause the difference in environment.

This will be shown in our data.
Organisms in higher abundance in the media samples compared to the environmental samples are most likely caused by selection pressures the media introduces.
To keep our data manageable in a workshop we will look at the phyla abundances.
If a phyla is detected to have higher abundance in a media compared to the environmental sample, we can say the phyla is a biomarker of the media compared to the environmental sample.
Additionally, we will be able to detect phyla that have been depleted in the media samples versus the environmental sample.

## DA: setup

We will use a new notebook called **"7-differential_abundance.ipynb"**. 
Add the following to the top of this new notebook to load the required libraries/packages and data.

```{r, eval=FALSE}
#Libraries
library("phyloseq")
library("microbiome")
library("IRdisplay")
library("ANCOMBC")
library("DT")
library("dplyr")
library("tidyr")
#Load phyloseq object
load("phyloseq.RData")
```

We will be using the package [`ANCOMBC`](https://github.com/FrederickHuangLin/ANCOM-BC-Code-Archive) to carry out our DA analysis.
Instead of using our rarefied data we will use our original abundance data.
This is because ANCOM-BC uses absolute abundances and corrects for differences between samples.
This is a statistically heavy method and I recommend you read the [ANCOM-BC paper](https://www.nature.com/articles/s41467-020-17041-7) if you are interested in how it works.
As bioinformaticians we are more interested in the output rather than the method.

## DA: Preprocess

When carrying out ANCOM analysis you need to specify the reference group for the groupings you use.
The reference group will be the 1st level of the factor.
Therefore we will check our level orders for our media and site metadata categories.

`r hide("What are factors?")`
For an explanation on the factors class please see the following section in the R primer omics book:
[Factors](https://neof-workshops.github.io/R_j4c0xh/14-Plots_scatterplot_and_box_plots.html#factors)
`r unhide()`

```{r, eval = FALSE}
#Check media and site levels
levels(phyloseq::sample_data(pseq)$media)
levels(phyloseq::sample_data(pseq)$site)
```

The levels are not in the order we would like.
We'll therefore change the order with the `factor()` function and `levels = ` option.

```{r, eval = FALSE}
#Ensure ENV is our reference level for media
sample_data(pseq)$media <- factor(sample_data(pseq)$media,
                                  levels = c("ENV", "CVP", "KBC", "TSA"))
levels(phyloseq::sample_data(pseq)$media)
#Ensure UD is our reference level for site
sample_data(pseq)$site <- factor(sample_data(pseq)$site,
                                  levels = c("UD", "MD", "LD"))
levels(phyloseq::sample_data(pseq)$site)
```

## DA: ANCOM-BC analysis

Now that we have our data the way we would like it we can carry out.
We'll use the function `ANCOMBC::ancombc2()` for our ANCOM-BC analysis.

__Note:__ This command will take a few minutes to run.

```{r, eval=FALSE}
#Run ANCOMBC
ancom_class_out <- ANCOMBC::ancombc2(data = pseq, fix_formula = "media + site",
                                     p_adj_method = "holm",
                                     tax_level = "Class")
```

The resulting object is very large so it can be a good idea to save it.

```{r, eval=FALSE}
#Save results
save(ancom_class_out, file = "ANCOMBC2_class_analysis.RData")
```

### ANCOMBC2 options

The options we use in the function are:

- `data = `: The data to be analysed. This can be provided in various types of objects including the `phyloseq` object.
- `fix_formula = `: The groupings within your data. It is importnat to add all groupings that may have an influence on the microbial abundances.
- `p_adj_method =`: The p adjustment value to use.
- `tax_level =`: Te taxonomic level of interest.

There are many more options that have been left as the default.
Please see the package [PDF](https://bioc.ism.ac.jp/packages/devel/bioc/manuals/ANCOMBC/man/ANCOMBC.pdf) for more options and explanations.

## DA: ANCOM-BC results

There is a lot of information in `ancom_class_out`.
Therefore, lets extract the results we want to look at and then remove the large object.

```{r, eval=FALSE}
#Output of analysis compared to reference groups
rec_ancom_class <- ancom_class_out£res
#Remove large object
rm(ancom_class_out)
#Check top rows of results
head(rec_ancom_class)
```

These reuslts only contain taxon where a significant result was found in at least one comparison.
This is based on the adjusted p-values (<0.05).
There are a lot of columns (37) in our result data frame.
Apart from taxon they are split by the comparison group.
These contain the following information:

- __taxon__: The taxa groups found to be a biomarker in at least one comparison.
- __lfc_*__: This contains the log (natural log) fold changes with respect to the reference group.
  - For example The values for lfc_mediaCVP would equal LFC(CVP - ENV).
  - A positive value means the abundances are higher in the comparison group (CVP) and lower in the ref (ENV).
  - A negative value means the abundances are lower in the comparison group (CVP) and higher in the ref (ENV).
  - The intercept value is the grand mean and is likely not of interest.
- __se_*__: The standard error for the LFC values.
- __W_*__: The W statistic. The higher this value the more significant the feature is differentially abundant between groups.
- __p_*__: The P-value.
- __q_*__: The Q-value. This is the adjusted P-value.
- __diff*__: This indicates whether there is a significant difference in the abundances of the comparison and reference group. In other words, is the Q-value < 0.05.

It is quite a large table so let's save it as a tsv (tab separated value) file.

```{r, eval=FALSE}
#Save as file
write.table(x = rec_ancom_class, file = "ANCOMBC2_class_results.tsv",
            quote = FALSE, row.names = FALSE)
```

This is good to open in excel but we don't have something like that on our cluster.
Therefore, we will use the package `DT` to create an interactive table with the function `datatable()`.
We can then save this object as an HTML with `htmlwidgets::saveWidget()`.

```{r, eval=FALSE}
#Create data table
m <- DT::datatable(res_ancom_class)
#Save data table as html
htmlwidgets::saveWidget(m, "ANCOM_class_results.html", selfcontained = FALSE)
```

Go to the jupyter-notebook file explorer tab and click on the html file to open it in a new tab.
This gives a relatively nice way to view all the results.

One very clear pattern is that there are no biomarkers found for MD and LD.
I believe this is because we don't have enough replicates for the sites within the media samples and the differences between the medias are so large.
However, there is a good amount of biomarkers for the three different medias compared to ENV.

## DA: Create signifcant LFC long data frame

The results table is good but let's make a heatmap showing the LFCs (Log fold changes) of the significant biomarkers. First, we need to do some preprecessing.

Remove columns unrelated to the media comparisons.

```{r, eval = FALSE}
#Extract the taxon column and the media columns from the results
#get positions of columns which contain "media"
media_colnames_pos <- grep("media", colnames(res_ancom_class))
#Taxon column and media column names
media_colnames <- colnames(res_ancom_class)[c(1,media_colnames_pos)]
media_colnames
#Create new df with only taxon and media columns
res_ancom_class_media <- media_colnames[,media_colnames]
head(res_ancom_class_media)
```

Because we had comparisons with more that 1 grouping (media & site) we might have rows with no significnat results.
We will therefore remove these rows.

__Note:__ The next 2 code sections use the packages `dplyr` & `tidyr` along with piping `%>%`.
This is used as it would be very difficult to do these steps in base R and they are the code provided in the [`ANCOMBC2` tutorial](https://bioconductor.org/packages/release/bioc/vignettes/ANCOMBC/inst/doc/ANCOMBC2.html).
If you would like to learn more about these packages please see the [R for Data Science book](https://r4ds.had.co.nz/).

```{r, eval = FALSE}
#Remove any taxon where there are no TRUE values
#I.e. no significant difference found for the media samples
#Number of biomarkers
nrow(res_ancom_class_media)
#Extract rows that contain at least one TRUE value
res_ancom_class_media_sig <- res_ancom_class_media %>% dplyr::filter_all(
  dplyr::any_vars(. %in% TRUE)
)
#Number of rows and the top of the new df
nrow(res_ancom_class_media_sig)
head(res_ancom_class_media_sig)
```

Next we need to do a variety of preprocessing steps.
This will create an edited long data frame.
Long data frames are needed for `ggplot2()` when not using a `phyloseq` object.
Instead of having the LFC (log fold change) values over many columns there will be only one column for the values. 
The long data frame will contain the following columns:

- __taxon__: The biomarker name.
- __group__: The comparison group.
- __value__: The LFC value.

Please see the code and its annotation below.
If you do not fully understand the code that is fine.
As long as you can edit it to work with your own future code.

```{r, eval = FALSE}
#Create edited long df to plot significant log fold changes on heatmap
res_ancom_class_media_sig_lfc_fig <- res_ancom_class_media_sig %>%
  #Change log change values to 0 if diff is false
  #test: Is diff TRUE?
  #yes: Set LFC to original value (e.g. lfc_mediaCVP)
  #no: Set LFC to 0
  dplyr::mutate(lfc_mediaCVP = ifelse(test = diff_mediaCVP == TRUE,
                                      yes = lfc_mediaCVP, no = 0),
                lfc_mediaKBC = ifelse(test = lfc_mediaKBC == TRUE,
                                      yes = lfc_mediaKBC, no = 0),
                lfc_mediaTSA = ifelse(test = lfc_mediaTSA == TRUE,
                                      yes = lfc_mediaTSA, no = 0)) %>%
  #Only retain taxon and log fold changes columns
  #Additionally, change column names and round LFC values to 2 decimal places
  dplyr::transmute(taxon,
                   CVP = round(lfc_mediaCVP,2),
                   KBC = round(lfc_mediaKBC,2),
                   TSA = round(lfc_mediaTSA,2)) %>%
  #Convert to long data frame
  tidyr::pivot_longer(cols = CVP:TSA,
                      names_to = "group",
                      values_to = "value") %>%
  #Arrange order of rows based on taxons
  dplyr::arrange(taxon)
#Check new long df
head(res_ancom_class_media_sig_lfc_fig)
```

## DA: Create signifcant LFC long data frame

ANCOM-BC (https://www.bioconductor.org/packages/release/bioc/vignettes/ANCOMBC/inst/doc/ANCOMBC.html)
ANCOM-BC github repo (https://github.com/FrederickHuangLin/ANCOMBC)